---
title: "COVID-19 in the US: Trends and analysis using ggplot2 in R"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
runtime: shiny    
---

```{r setup, include=FALSE}
library(flexdashboard)
```

###### By: Zihua Lai & Daniela Quintero Narváez.

**Introduction**

The project aims to visualize the state-wise COVID-19 data in the United States, focusing on trends, comparisons, and correlations within the data. We will use the ggplot2 package in R for exploring and visualizing the data, which will result into meaningful insights about the pandemic's impact in the country and at the state level.

The data was obtained from the official website of The COVID tracking project: <https://covidtracking.com/data/download>

**Objectives:**

1.  To explore and visualize state-wise trends across the country regarding COVID-19 cases, deaths, hospitalizations, and testing.

2.  To compare the pandemic's impact across different states of the country.

3.  To explore the correlations between various indicators such as positive cases, testing rates, and hospitalizations.

4.  To identify patterns and outliers in the data.

**Data-set description**

| Variable                           | Description                                                                                                                                                                                                                                                                                    |
|---------------------------|-------------------------------------------------------------------------------------------------|
| `date`                             | Date on which data was collected by The COVID Tracking Project.                                                                                                                                                                                                                                |
| `state`                            | Two-letter abbreviation for the state or territory.                                                                                                                                                                                                                                            |
| `death`                            | Daily increase in *hospitalizedCumulative*, calculated from the previous day's value.                                                                                                                                                                                                          |
| `deathConfirmed`                   | Total **fatalities with confirmed COVID-19 case diagnosis.**                                                                                                                                                                                                                                   |
| `deathIncrease`                    | Daily increase in death, calculated from the previous day's value.                                                                                                                                                                                                                             |
| `deathProbable`                    | Total **fatalities with probable COVID-19 case diagnosis**                                                                                                                                                                                                                                     |
| `hospitalized`                     | Total number of **unique** **individuals who have** **ever been hospitalized** with COVID-19.                                                                                                                                                                                                  |
| `hospitalizedCumulative`           | Total number of individuals who have **ever been hospitalized with COVID-19**.                                                                                                                                                                                                                 |
| `hospitalizedCurrently`            | Individuals who are **currently hospitalized with COVID-19.**                                                                                                                                                                                                                                  |
| `hospitalizedIncrease`             | Daily increase in *hospitalizedCumulative*, calculated from the previous day's value.                                                                                                                                                                                                          |
| `inIcuCumulative`                  | Total number of individuals who have **ever been hospitalized in the Intensive Care Unit with COVID-19**                                                                                                                                                                                       |
| `inIcuCurrently`                   | Individuals who are **currently hospitalized in the Intensive Care Unit with COVID-19**                                                                                                                                                                                                        |
| `negative`                         | Total number of **unique people with a completed PCR test that returns negative**.                                                                                                                                                                                                             |
| `negativeIncrease`                 | Daily increase on negative variable                                                                                                                                                                                                                                                            |
| `negativeTestsAntibody`            | The total number of completed antibody tests that return negative as reported by the state or territory.                                                                                                                                                                                       |
| `negativeTestsPeopleAntibody`      | The total number of unique people with completed antibody tests that return negative as reported by the state or territory.                                                                                                                                                                    |
| `negativeTestsViral`               | Total number of completed PCR tests (or specimens tested) that return negative as reported by the state or territory                                                                                                                                                                           |
| `onVentilatorCumulative`           | Total number of individuals who have **ever been hospitalized under advanced ventilation with COVID-19.**                                                                                                                                                                                      |
| `onVentilatorCurrently`            | Individuals who are **currently hospitalized under advanced ventilation with COVID-19.**                                                                                                                                                                                                       |
| `positive`                         | Total number of confirmed plus probable cases of COVID-19 reported by the state or territory                                                                                                                                                                                                   |
| `positiveCasesViral`               | Total number of unique people with a positive PCR or other approved nucleic acid amplification test (NAAT), as reported by the state or territory.                                                                                                                                             |
| `postiveIncrease`                  | The daily increase in field *positive*, which measures cases(confirmed + probable) calculated based on the previous day's value.                                                                                                                                                               |
| ~~`positiveScore`~~                | This variable is not defined in the website but also all value are 0, so we will drop it from the project.                                                                                                                                                                                     |
| `positiveTestsAntibody`            | Total number of completed antibody tests that return positive as reported by the state or territory.                                                                                                                                                                                           |
| `positiveTestsAntigen`             | Total number of **completed antigen tests that return positive** as reported by the state or territory                                                                                                                                                                                         |
| `positiveTestsPeopleAntibody`      | The total number of unique people with completed antibody tests that return positive as reported by the state or territory.                                                                                                                                                                    |
| `positiveTestsPeopleAntigen`       | Total number of **unique people with a completed antigen test that returned positive** as reported by the state or territory                                                                                                                                                                   |
| `positiveTestsViral`               | Total number of completed PCR tests (or specimens tested) that return positive as reported by the state or territory                                                                                                                                                                           |
| `recovered`                        | Total number of **people that are identified as recovered from COVID-19**. Types of "recovered" cases include those who are discharged from hospitals, released from isolation, or those who have not been identified as fatalities after a number of days (30 or more) post disease onset.    |
| `totalTestEncountersViral`         | Total number of people tested *per day* via PCR testing as reported by the state or territory.                                                                                                                                                                                                 |
| `totalTestEncountersViralIncrease` | Daily increase in totalTestEncountersViral                                                                                                                                                                                                                                                     |
| `totalTestResults`                 | In most states, the totalTestResults field is currently computed by adding positive and negative values because, historically, some states do not report totals, and to work around different reporting cadences for cases and tests.                                                          |
| ~~`totalTestResultsIncrease`~~     | Daily increase in *totalTestResults*, calculated from the previous day's value. This calculation includes all the caveats associated with Total tests/*totalTestResults*, and **it is recommended against using it at the state/territory level**, so we will not be using it in this project. |
| `totalTestsAntibody`               | Total number of **completed antibody tests** as reported by the state or territory                                                                                                                                                                                                             |
| `totalTestAntigen`                 | Total number of **completed antigen tests**, as reported by the state or territory.                                                                                                                                                                                                            |
| `totalTestsPeopleAntibody`         | The total number of **unique people who have been tested at least once via antibody testing** as reported by the state or territory.                                                                                                                                                           |
| `totalTestsPeopleAntigen`          | Total number of **unique people who have been tested at least once via antigen testing**, as reported by the state or territory                                                                                                                                                                |
| `totalTestsPeopleViral`            | Total number of unique people tested at least once via PCR testing, as reported by the state or territory.                                                                                                                                                                                     |
| `totalTestsPeopleViralIncrease`    | Daily increase in totalTestsPeopleViral                                                                                                                                                                                                                                                        |
| `totalTestsViral`                  | Total number of PCR tests (or specimens tested) as reported by the state or territory.                                                                                                                                                                                                         |
| `totalTestsViralIncrease`          | Daily increase in totalTestsViral                                                                                                                                                                                                                                                              |

> ##### Data-set definition sourced from the CoVid-tracking website: <https://covidtracking.com/about-data/data-definitions>

```{r install_packages, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
packages_to_install <- c("rmarkdown", "renv", "shiny", "ggplot2", "dplyr", "tidyr", "usmap")
install.packages(packages_to_install)
```

### **Data cleaning and exploratory analysis**

```{r libraries, message=FALSE, warning=FALSE, paged.print=TRUE, fig.height=20}
library(renv)
library(shiny)
library(ggplot2)
library(dplyr)
library(tidyr)
library(usmap)

data <- read.csv("data/all-states-history (1).csv",sep=",")
print(summary(data))
```

The summary of the data shows significant missing values for some variables, but due to the complexity of the attributes and the way the data was constructed we will replace them with 0.

Additionally, it is necessary to categorize the variables properly before moving to the visualization and analysis:

```{r, message=FALSE, warning=FALSE}
str(data)

```

We need to categorize `date` as a date class and `state` as factor to be able to visualize the data properly in the plots and in the dashboard page. We will also drop the column `totalTestResultsIncrease` as it was recommended not to use it by the creators of the data due to inaccuracies.

```{r dataTransform, message=FALSE, warning=FALSE}
data_transformed <- data
data_transformed$date <- as.Date(data_transformed$date)
data_transformed$state <- as.factor(data_transformed$state)

data_transformed <- replace(data_transformed, is.na(data_transformed), 0)

data_transformed <- data_transformed %>%
  select(-c("positiveScore", "totalTestResultsIncrease"))
```

After transformation, the data:

```{r}
str(data_transformed)
summary(data_transformed)
```

Variable `date` is now as in date format, which will allow us to graph the time series in a better way. Also, `states` are now as factor, with 56 levels; which consist of the 50 states, the federal district and 5 major territories (American Samoa, Guam, Northern Mariana Islands, Puerto Rico and the US virgin islands).

### Visualizing the data

*All the plots presented below are in the second page called Dashboard. You can go directly there and play with the graphs as it was built in as an interactive shiny dashboard. For details about the construction of the plots and code chunks, please keep reading below.*

[Here would be nice to plot the USA map by states and show with colors variables like confirmed deaths, positive cases, total tests done, etc. So far I was trying to map this but didnt manage :( Maybe you can try with your dear friend GPT 4.0?]{.underline}

[Some statistics about which state had the most cases, deaths, people in ICU, recovered]{.underline}

[Then we can plot histograms for the variables, maybe by state? there are 56 states data, maybe 7x8 grid with histograms or box plots??]{.underline}

And then we can use the paper you sent me and plot nicely other variables :)

If we have sometime we can build a dashboard to get more points, if not, we can forget about the part below :)

# Dashboard {data-icon="fa-signal"}

```{r global, include=FALSE}
d <- data_transformed
```

### USA Plot

## Sidebar {.sidebar}

Here we can put to select states (if we want to visualize data for only one state or many) and also which variable to visualize.

```{r}
selectInput(
  "n_breaks", label = "Number of bins:",
  choices = c(10, 20, 35, 50), selected = 20
)

sliderInput(
  "bw_adjust", label = "Bandwidth adjustment:",
  min = 0.2, max = 2, value = 1, step = 0.2
)
```

## Plot/Plots

the code is from an example I saw on internet :) just as a reference

```{r}
renderPlot({
  erpt = faithful$eruptions
  hist(
    erpt, probability = TRUE, breaks = as.integer(input$n_breaks),
    xlab = "Duration (minutes)", main = "Geyser Eruption Duration",
    col = 'gray', border = 'white'
  )
  
  dens = density(erpt, adjust = input$bw_adjust)
  lines(dens, col = "blue", lwd = 2)
})
```

```{r}


```
